---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include = FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Summarizing Data

Once you have collected the data for your experiment, how do you start
analyzing the data? A good first step is to look at the summary statistics
and the central measures. 

Assume we have collected data on the effect of two soporific drugs on 10
different patients. We want to see if these two drugs do indeed increase
patients' hours of sleep per night. The data is in a variable called `sleep`.

The data has three variables: `extra`, `group`, and `ID`. 

`extra` represents
the change in the amount of hours of sleep per night. A positive number here
represents an increase in hours of sleep compared to the control. A negative 
number represents a decrease in hours of sleep compared to the control. 

`group` represents the type of drug given (drug 1 or drug 2).

`ID` represents the patient ID. 

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

We want to first look at the data we collected. We can look at the
first couple of rows of our data set by calling the `head()` function. 

```{r head, exercise = TRUE}
head(sleep)
```

If you get an error that `sleep` does not exist, then load 
`library(datasets)`. You can also call sleep straight from its namespace as 
follows.

```{r namespace, exercise = TRUE}
head(datasets::sleep)
```

### Exercise

The drug used is currently displayed as a number. How can we know if it is
coded as a numeric or a factor in the data frame? To do this, use
the function `summary()` on the data set `sleep`. `summary()` is a very
useful function because it can provide a quick overview of both
numeric and factor variables. For numeric values it 

```{r summary, exercise = TRUE}
summary()
```

```{r types}
quiz(
  question("How are the three variables typcasted in the sleep data?",
    answer("extra = numeric, group = factor, ID = factor", correct = TRUE),
    answer("extra = factor, group = factor, ID = factor"),
    answer("extra = numeric, group = numeric, ID = factor"),
    answer("extra = factor, group = factor, ID = numeric"))
  )
```

```{r typecast}
quiz(
  question("Why is it important that we know how a variable is casted in the dataframe?",
    answer("It's not important since all variables in a dataframe are the same type."),
    answer("We can convert variables to the type we want to make analysis easier.", correct = TRUE),
    answer("We can use the appropriate functions for different types of variables.", correct = TRUE),
    answer("We can rename the variables to a name that suits its type."))
  )
```

We see that the mean increase in hours of sleep for both types of drugs is
`r mean(sleep$extra)`. However, we have patients that have had maximum
increase of `r max(sleep$extra)` hours of sleep and a loss of 
`r min(sleep$extra)` hours of sleep. This leads to the question - is 
one drug more effective than the other?

### Exercise 

Let's look at the summary statistics of the entries where patients took drug 1 
and compare it to the overall distribution. 

```{r print-limit, exercise = TRUE, exercise.eval = TRUE}
summary(sleep)
```

```{r print-limit-hint}
# Note that we have to first subset our data such that the observations where the
# patients took drug 1 are being kept 

sleep$group =="1"

# We subset a dataframe the same way we subset a matrix
# We want to subset the rows, not the columns in this case

sleep[4,]
```

```{r control}
quiz(
  question("True or False: Patients who took drug 1 sleep more than the control.",
    answer("True", correct = TRUE),
    answer("False"))
  )
```

