---
title: "R Basics I: Starting New Projects"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include = FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```

## Goals of Tutorial

Starting a new project can be daunting and even scarier than writing your 
RBA paper for PWR. This tutorial will guide you through good practices when 
first starting to work with data. 

Goals for this session include:

1. Looking for help in RStudio;
2. Understanding the data and data frame using exploratory data analysis;
3. Examining what can go wrong when performing simple tasks in R. 

For this tutorial, assume we have collected data on the effect of 
two soporific drugs (drugs that make you sleepy) on 10
different patients. We want to see if these two drugs do indeed 
increase patients' hours of sleep per night. 

The data is in a data frame called `sleep`.

The data has three variables: `extra`, `group`, and `ID`. 

`extra` represents
the change in the amount of hours of sleep per night. A positive number here
represents an increase in hours of sleep compared to the control. A negative 
number represents a decrease in hours of sleep compared to the control. 

`group` represents the type of drug given (drug 1 or drug 2).

`ID` represents the patient ID. 

Let's get started! Click on "Next Topic" to begin.

## Understanding Data

Once you have collected the data for your experiment, how do you start
analyzing the data? A good first step is to first understand where the data 
came from. If you are doing your own research, then you should know the ins
and outs. But what if you were handed a dataset from a specific R package?

You can use `?sleep` to understand it more. You can use `?` for a function,
dataset, or anything you don't understand in R to get help. Make sure you
are connected to the internet. A help page should appear in one of your 
panels in RStudio. 

```{r help, exercise = TRUE}
?sleep
```

Let's first look at the first few rows of our `sleep` data. Use the `head()`
function. You can use `?head` for examples and details about the function. 

```{r head, exercise = TRUE}
?head
```

Now use head on the dataset `sleep`. 

```{r head_in, exercise = TRUE}
head()
```

If you get an error that `sleep` does not exist, then load 
`library(datasets)`. You can also call sleep straight from its namespace as 
follows. But note you might have to continue calling it straight from its
namespace for the rest of the exercises.

This technique is particularly useful when you have two package with the
same function. By specifying the package the function is from, R won't get
confused and throw you unnecessary errors. 

```{r namespace, exercise = TRUE}
head(datasets::sleep)
```

Note that at a first glance, all of our values are numeric. It looks like 
there are three variables in our
data. We can figure out the full dimension of our data by using the `dim()`
function.

```{r dim, exercise = TRUE}
dim()
```

We see the function returns 2 numbers. The first number is how 
many rows are in the data frame. The second number is how many columns there are
in the data frame. If you want to know just the number of rows or columns, you can 
use `nrow(sleep)` or `ncol(sleep)`. 

```{r row, exercise = TRUE}
nrow()
```

There are 20 rows in our data frame.

```{r col, exercise = TRUE}
ncol()
```

There are 3 columns in our data frame. 

## Structure of Data

Now that we know what our data's physical attributes are, we want to further
understand what each variable's type is. By understanding variable type,
we can further determine what kinds of operations we can perform with
different types of variables. Furthermore, variable types are especially
important when formualating your analysis attack plan for your project.
Categorical variables and numeric variables require ifferent models and 
different visualization techniques.

### Exercise

The drug used is currently displayed as a number. How can we know if it is
coded as a numeric or a factor in the data frame? To do this, use
the function `summary()` on the data set `sleep`. 

`summary()` is a very
useful function because it can provide a quick overview of both
numeric and factor variables. For numeric values it gives us a quantile summary.
For factors it gives us a frequency count. 

```{r summary, exercise = TRUE}
summary()
```

You can also use `str()` (stands for "structure") if you just want a 
quick glimpse of the data and not a full-blown summary.

```{r structure, exercise = TRUE}
str()
```

```{r types}
quiz(
  question("How are the three variables typcasted in the sleep data?",
    answer("extra = numeric, group = factor, ID = factor", correct = TRUE),
    answer("extra = factor, group = factor, ID = factor"),
    answer("extra = numeric, group = numeric, ID = factor"),
    answer("extra = factor, group = factor, ID = numeric"))
  )
```

```{r typecast}
quiz(
  question("Why is it important that we know how a variable is casted in the dataframe?",
    answer("It's not important since all variables in a dataframe are the same type."),
    answer("We can convert variables to the type we want to make analysis easier.", correct = TRUE),
    answer("We can use the appropriate functions and techniques for different types of variables.", correct = TRUE),
    answer("We can rename the variables to a name that suits its type."))
  )
```

We see that the mean increase in hours of sleep for both types of drugs is
`r mean(sleep$extra)`. However, we have patients that have had maximum
increase of `r max(sleep$extra)` hours of sleep and a loss of 
`r min(sleep$extra)` hours of sleep. This leads to the question - is 
one drug more effective than the other?

## Visualizing Data

In order to answer whether one drug is more effective in inducing sleep
than the other, we can start by looking at some numbers. 

### Exercise 

Let's look at the summary statistics of the entries where patients took drug 1 
and compare it to the overall distribution. You can use `filter()` to only
look at the observations where `group == 1`. Note that 1 is not a numeric
in the data set, so we have to put it in quotes.

```{r print-limit, exercise = TRUE, exercise.eval = TRUE}
sleep %>% 
  filter()
```

The piping function `%>%` is a tool that brings the data from the first 
variable call to the function. This allows us to type `sleep %>% filter(extra > 0)`
as opposed to nesting all of the functions together in Base R 
`sleep[sleep$extra > 0,]`. Note that we also don't need `filter(sleep$extra > 0)` 
since the `sleep` data frame is piped in. 

The piping function is useful because it will also allow us to build
layers into our data manipulation and data visualization code. 

```{r print-limit-hint}
# After filtering for the group you want, you need to pipe it into the summary()
# function. Since the summary() function receives the argument from the previous pipe, 
# it does not need an argument

sleep %>% 
  filter() %>% 
  summary()
```

```{r control}
quiz(
  question("True or False: Patients who take drug 1 sleep more than the control.",
    answer("True", correct = TRUE),
    answer("False"))
  )
```

It is hard to interpret numbers without starting at them until your eyes hurt.
Let's try to visualize what the data is telling us. 

We want to understand the overall ranges of the different types of drugs.
To visualize range, we can use a boxplot. Try plotting using `ggplot()`.
While `ggplot()` is also in the tidyverse, note that its connector is not the
pipe %>% but the +. The `+` adds layer to your picture, which is established
by a blank canvas call with `ggplot()`. `geom_boxplot()` will call a
boxplot layer for you. The `mapping` argument will allow you to places
variables where you want them to go (on the x and y-axis in this case). Use
`?geom_boxplot()` to learn more. 

```{r print-boxplot, exercise = TRUE}
sleep %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = , y = ))
```

```{r print-boxplot-hint}
# Your x value is typically a factor. What do we want to distinguish the two groups 
# by? Perhaps by drug (group)?
```

```{r results}
quiz(
  question("Does drug 1, on average, make patients sleepier or less sleepy than those on drug 2?",
    answer("More sleepy"),
    answer("Less sleepy", correct = TRUE)
  )
)
```

Since drug 1 on average makes patients less sleepy, we did notice the whiskers
on the boxplot. How large exactly is the variance on `extra`? To examine the
distribution, make a histogram with `geom_histogram` to explore this question.
Remember to adjust the binwidth!

```{r print-hist, exercise = TRUE}
sleep %>% 
  filter() %>% 
  ggplot() +
  geom_histogram(mapping = aes(x = ), binwidth = )
```

```{r print-hist-hint}
# Try different binwidths to see how the histogram changes. Which one is the most
# informative?
```

Assume that the `ID` value is actually not different individuals, but actually
a measure of a user's age over time. We are curious as to how drug 1 affected
a patient over time. How can we visualize trends over time or with a certain order?
With a line plot of course!

There are 3 steps to this exercise. We have to first typecast the `ID` 
variable into a numeric. Use `mutate()` and the piping connector to do this.
Then we want to overlay `geom_point` to mark our
points on the graph first. This is how you create a scatter plot. Then to
connect the points, we overlay `geom_line`. Try it out in the exercise. This
one's a little tricky, so if you need help try the hint. 

```{r print-point, exercise = TRUE}
sleep %>% 
  ggplot() +
  geom_point() +
  geom_line()
```

```{r print-point-hint}
sleep %>% 
  filter(group == "1") %>% 
  mutate(ID = as.numeric(ID)) %>% 
  # We will be using the same x and y arguments for 2 geoms, so it is better to put 
  # it in the general ggplot() call
  ggplot(mapping = aes(x = , y = )) +
  geom_point() +
  geom_line()
```

Are there any particular trends you see with the data? For a specific individual,
we see that there is an extra 2 - 3 hours of sleep near the 6 or 7th index. 
However, the sleep returns to similar levels as the 1st or 2nd index. If we had
more data, it is possible that the pattern is more cyclical. 

Again, this is
only assuming that ID represents a time measure rather than individuals! If you
plot a factor on the x-axis, then the graph has no real meaning (because we 
can't really see any trends). 

## Data Manipulation

In the instance where you have extra information regarding your data or if
you measured your observations using auxiliary information, you may want to 
alter the values in your data set and save them. This requires variable
assignment (you can get more practice with DataCamp) and `mutate()`, 
`filter()` which we used above. 

The hint will give you an answer. So take a peak to see if you wrote 
it correctly.

### Exercises

We were given extra information about the 5th observation in our data frame.
We want to look only at the 5th row in our data frame. How do we index
the 5th row?

```{r index, exercise = TRUE}
sleep
```

```{r index-hint}
sleep[5,]
```

We want to use the `extra` variable of the 5th observation as a parameter in
another calculation we are doing. How do we extract this value? How do we
index the `extra` variable of the 5th observation?

```{r rowdex, exercise = TRUE}
sleep
```

```{r rowdex-hint}
sleep[5, 1]
```

Next, we realized that we might have made a mistake in our experiment set up
for drug group 2. We want to change all of the extra values in group 2 to be 
0.5 less than they are because we forgot to start the timer on time. 

```{r manip, exercise = TRUE}
sleep %>% 
  filter()
```

```{r manip-hint}
sleep %>% 
  filter(group == "2") %>% 
  mutate(extra = extra - 0.5)
```

Since we have 20 observations, that is a lot of data to put in a presentation.
We want a condensed table with the averages for each group of drug tested.
The format of the table should give me two variables (average_extra and group).
Where `average_extra` is the average of all the extra values for that
specific group number. 

This one uses `group_by()`, which is essentially breaking down the data set
by a variable and then performing the requested operation. `group_by()`
performs a function similar to `filter()` in the sense that it filters the
data set many times such that for each value of the variable we are grouping by,
we can `summarise()` the data for that group. 

Both `summarize()` and `summarise()` are the same. The founder of dplyr is from 
New Zealand, so he wanted to be inclusive.

The function to take the average is `mean()`. 

```{r group, exercise = TRUE}
sleep %>% 
  group_by() %>% 
  summarise()
```

```{r group-hint}
sleep %>% 
  group_by(group) %>% 
  summarise(extra = mean(extra))
```

Now let's look at the sum of the total quantity of sleep hours lost. This means
negative `extra` should be counted as positive. You can use the `abs()` function
to take absolute value.

```{r abs, exercise = TRUE}
sleep %>% 
  group_by() %>% 
  summarise()
```

```{r abs-hint}
sleep %>% 
  group_by(group) %>% 
  summarise(extra = extra %>% abs() %>% sum())
```

Hope you all got a good grasp on basic data wrangling. If you want to learn
more you can go to this website ([R for Data Science by Hadley Wickham and
Garrett Grolemund](http://r4ds.had.co.nz/)).

## Final Thoughts

Congratulations! 

You just walked through the beginning of a project. Now what
drives your project is your curiosity and your ability to ask good questions.
This is the part of data work that is most like a detective novel. You should
be applying what you learned in class to sharpen your intuition. A key
component of good data work is asking the right questions and knowing the 
right tools to answer those questions. 

Now you can procrastinate! Yay!
